\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-04-28 submission/manuscript.Rmd
Final edits and cleaning

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had healthy colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals, of which 52 had an SRN (54%). The training data were used to build and select the models, and the test set was used for evaluating the model. We trained seven different models using the training data [Table 1].

@@

**Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to calculate decision values for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel. The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.

@@

\small
\begin{tabular}{|l|l|l|}
\hline

\rowcolor{lightgray}
\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline

\makecell[l]{Logistic regression} & \makecell[l]{A predictive regression analysis when the dependent variable is binary.} & Linear \\ \hline

\makecell[l]{SVM with linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear separating hyperplane that discriminates between labels.} & Linear  \\ \hline

\makecell[l]{SVM with radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear separating hyperplane that discriminates between labels.} & Non-linear \\ \hline

\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
root to the leaf node where an attribute is tested to discriminate between labels.} & Non-linear \\ \hline

Random forest & \makecell[l]{A classifier that is an ensemble of decision trees that grows randomly with subsampled data.} & Non-linear \\ \hline

\makecell[l]{Gradient Boosted Trees (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees that grows greedily.} & Non-linear \\ \hline

\end{tabular}
\newpage

@@

>>>>

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had healthy colons (53.3%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 184 had an SRN (46.8%), while the test set was composed of 97 individuals, of which 45 had an SRN (46.4%). The training data were used to build and select the models, and the test set was used for evaluating the model. We trained seven different models using the training data [Table 1].

@@

**Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to calculate decision values for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel. The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear4.R}, respectively.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/}.

@@

| **Model** | **Description** | **Linearity** |
|-----------|-----------------|---------------|
|  Logistic regression |  A predictive regression analysis when the dependent variable is binary | Linear  |
|  SVM with linear kernel | A classifier that is defined by an optimal linear separating hyperplane that discriminates between labels  | Linear  |
|  SVM with radial basis kernel | A classifier that is defined by an optimal non-linear separating hyperplane that discriminates between labels  | Non-linear  |
|  Decision tree | A classifier that sorts samples down from the root to the leaf node where an attribute is tested to discriminate between labels  | Non-linear  |
|  Random forest | A classifier that is an ensemble of decision trees that grows randomly with subsampled data  | Non-linear |
|  Gradient Boosted Trees (XGBoost) | A classifier that is an ensembe of decision trees that grows greedily  | Non-linear  |

@@

**Table S1. An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.**

### 2020-04-28 submission/manuscript.tex
Final edits and cleaning

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or screen relevant
neoplasia using Bray-Curtis distances based on the 16S rRNA gene
sequences collected from fecal samples {[}Figure S1{]}. This variation
is likely due to the ability of many bacterial populations to fill the
same niche such that different populations cause the same disease in
different individuals. Furthermore, a growing number of studies have
shown that it is rare for a single bacterial species to be associated
with a disease. Instead, subsets of the microbiome account for
differences in health. Traditional statistical approaches do not
adequately account for the variation in the human microbiome and
typically consider the protective or risk effects of each bacterial
population separately (1). Recently, machine learning (ML) models have
grown in popularity among microbiome researchers as our ability to
sample large numbers of individuals has grown; such models can
effectively account for the interpersonal microbiome variation and the
ecology of disease because they consider the relative abundance of each
bacterial population in the context of others rather than in isolation.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had healthy colons (53\%) and 229 had a
screen relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals, of which 52 had an SRN (54\%). The training
data were used to build and select the models, and the test set was used
for evaluating the model. We trained seven different models using the
training data {[}Table 1{]}.

@@

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data
{[}Figure S6{]}. We then calculated how much the predictive performance
of the model (i.e., testing AUROC values) decreased when each OTU or
group of OTUs was randomly permuted. We ranked the OTUs based on how
much the median testing AUROC decreased when it was permuted; the OTU
with the largest decrease ranked highest {[}Figure 4{]}. Among the
twenty OTUs with the largest impact, there was only one OTU (OTU 822)
that was shared among all of the models; however, we found three OTUs
(OTUs 58, 110, 367) that were important in each of the tree-based
models. Similarly, the random forest and XGBoost models shared four of
the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
results also revealed that with the exception of the decision tree
model, removal of any individual OTU had minimal impact on model
performance. For example, if OTU 367 was permuted across the samples in
the decision tree model, the median AUROC dropped from 0.601 to 0.525.
In contrast, if the same OTU was permuted in the random forest model,
the AUROC only dropped from 0.695 to 0.680, which indicated high degree
of collinearity in the dataset. Permutation analysis allowed us to gauge
the importance of each OTU in non-linear models and partially account
for collinearity by grouping correlated OTUs to determine their impact
as a group.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figures S2-S3; Figure 5{]}.

@@

Our framework provides a reproducible structure for investigators
wanting to train, evaluate, and interpret their own ML models to
generate hypotheses regarding which OTUs might be biologically relevant.
However, deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(44). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (45). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values for models generated using L2-regularized SVM
with linear kernel and L1-regularized SVM with linear kernel. The code
for these changes on L2-regularized SVM with linear kernel and
L1-regularized SVM with linear kernel models are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
respectively.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.

@@

\captionsetup{labelformat=empty}
\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
\small
\begin{tabular}{|l|l|l|}
\hline

\rowcolor{lightgray}
\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline

\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline

\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline

\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline

\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline

Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline

\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline

\end{tabular}
\newpage

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross-validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by a
horizontal line at 0.50) for all the ML models. The predictive
performance of random forest model was higher than other ML models,
though not significantly (p \textgreater{} 0.05). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel performances were not significantly different from
one another. The boxplot shows quartiles at the box ends and the median
as the horizontal line in the box. The whiskers show the farthest points
that were not outliers. Outliers were defined as those data points that
are not within 1.5 times the interquartile ranges.

@@

\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
with radial basis kernel, (B) decision tree, (C) random forest, and (D)
XGBoost feature importances were explained using permutation importance
on the held-out test data set. The gray rectangle and the dashed line
show the IQR range and median of the base testing AUROC without any
permutation. The 20 OTUs that caused the largest decrease in the AUROC
when permuted are reported here. The colors of the box plots represent
the OTUs that were shared among the different models; yellow were OTUs
that were shared among all the non-linear models, green were OTUs that
were shared among the tree-based models, green were the OTUs shared
among SVM with radial basis kernel, decision tree and XGBoost, pink were
the OTUs shared among SVM with radial basis kernel and XGBoost only, red
were the OTUs shared among random forest and XGBoost only and blue were
the OTUs shared among decision tree and random forest only. For all of
the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
had the largest impact on predictive performance.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or screen relevant
neoplasia using Bray-Curtis distances based on the 16S rRNA gene
sequences collected from fecal samples {[}Figure S1{]}. This variation
is likely due to the ability of many bacterial populations to fill the
same niche such that different populations cause the same disease in
different individuals. Furthermore, a growing number of studies have
shown that it is rare for a single bacterial species to be associated
with a disease. Instead, subsets of the microbiome account for
differences in health. Traditional statistical approaches do not
adequately account for the variation in the human microbiome and
typically consider the protective or risk effects of each bacterial
population separately (1). Recently, machine learning (ML) models have
grown in popularity among microbiome researchers because ML models can
effectively account for the interpersonal microbiome variation and the
ecology of disease as they consider the relative abundance of each
bacterial population in the context of others rather than in isolation.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had healthy colons (53.3\%) and 229 had a
screen relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 184 had an SRN (46.8\%), while the test set was
composed of 97 individuals, of which 45 had an SRN (46.4\%). The
training data were used to build and select the models, and the test set
was used for evaluating the model. We trained seven different models
using the training data {[}Table 1{]}.

@@

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data
{[}Figure S6{]}. We then calculated how much the predictive performance
of the model (i.e., testing AUROC values) decreased when each OTU or
group of OTUs was randomly permuted. We ranked the OTUs based on how
much the median testing AUROC decreased when it was permuted; the OTU
with the largest decrease ranked highest {[}Figure 4{]}. Among the
twenty OTUs with the largest impact, there was only one OTU (OTU 822)
that was shared among all of the models. We also found that three OTUs
(OTUs 58, 110, 367) were important in each of the tree-based models.
Similarly, the random forest and XGBoost models shared four of the most
important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also
revealed that with the exception of the decision tree model, removal of
any individual OTU had minimal impact on model performance. For example,
if OTU 367 was permuted across the samples in the decision tree model,
the median AUROC dropped from 0.601 to 0.525. In contrast, if the same
OTU was permuted in the random forest model, the AUROC only dropped from
0.695 to 0.680, which indicated high degree of collinearity in the
dataset. Permutation analysis allowed us to gauge the importance of each
OTU in non-linear models and partially account for collinearity by
grouping correlated OTUs to determine their impact as a group.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figure 5{]}.

@@

Our framework provides a reproducible pipeline to train, evaluate, and
interpret microbiome-based ML models and generate hypotheses to explain
the underlying microbiology of the model prediction. However, deploying
microbiome-based models to make clinical diagnoses or predictions is a
significantly more challenging and distinct undertaking (44). For
example, we currently lack standardized methods to collect patient
samples, generate sequence data, and report clinical data. We are also
challenged by the practical constraints of OTU-based approaches. The de
novo algorithms commonly in use are slow, require considerable memory,
and result in different OTU assignments as new data are added (45).
Finally, we also need independent validation cohorts to test the
performance of a diagnostic model. To realize the potential for using ML
approaches with microbiome data, it is necessary that we direct our
efforts to overcome these challenges.

@@

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values for models generated using L2-regularized SVM
with linear kernel and L1-regularized SVM with linear kernel. The code
for these changes on L2-regularized SVM with linear kernel and
L1-regularized SVM with linear kernel models are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear4.R},
respectively.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/}.

@@

\textbf{Table 1.} Characteristics of the machine learning models in our
comparative study.

\begin{longtable}[]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.23\columnwidth}\raggedright
\textbf{Model}\strut
\end{minipage} & \begin{minipage}[b]{0.36\columnwidth}\raggedright
\textbf{Description}\strut
\end{minipage} & \begin{minipage}[b]{0.32\columnwidth}\raggedright
\textbf{Linearity}\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.23\columnwidth}\raggedright
Logistic regression\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A predictive regression analysis when the dependent variable is
binary\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Linear\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
SVM with linear kernel\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A classifier that is defined by an optimal linear separating hyperplane
that discriminates between labels\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Linear\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
SVM with radial basis kernel\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A classifier that is defined by an optimal non-linear separating
hyperplane that discriminates between labels\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Non-linear\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
Decision tree\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A classifier that sorts samples down from the root to the leaf node
where an attribute is tested to discriminate between labels\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Non-linear\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
Random forest\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A classifier that is an ensemble of decision trees that grows randomly
with subsampled data\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Non-linear\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
Gradient Boosted Trees (XGBoost)\strut
\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
A classifier that is an ensembe of decision trees that grows
greedily\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
Non-linear\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross-validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by a
vertical line at 0.50) for all the ML models. The predictive performance
of random forest model was higher than other ML models, though not
significantly (p \textgreater{} 0.05). L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel performances were not significantly different from one
another. The boxplot shows quartiles at the box ends and the median as
the horizontal line in the box. The whiskers show the farthest points
that were not outliers. Outliers were defined as those data points that
are not within 1.5 times the interquartile ranges.

@@

\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
with radial basis kernel, (B) decision tree, (C) random forest, and (D)
XGBoost feature importances were explained using permutation importance
on the held-out test data set. The gray rectangle and the dashed line
show the IQR range and median of the base testing AUROC without any
permutation. The 20 OTUs that caused the largest decrease in the AUROC
when permuted are reported here. The colors of the box plots represent
the OTUs that were shared among the different models; yellow were OTUs
that were shared among all the non-linear models, green were OTUs that
were shared among the tree-based models, turqoise were the OTUs shared
among SVM with radial basis kernel, decision tree and XGBoost, pink were
the OTUs shared among SVM with radial basis kernel and XGBoost only, red
were the OTUs shared among random forest and XGBoost only and blue were
the OTUs shared among decision tree and random forest only. For all of
the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
had the largest impact on predictive performance.

@@

\textbf{Table S1. An aspirational rubric for evaluating the rigor of ML
practices applied to microbiome data.}
